Avivo Demographics Markdown File
=====================================

The purpose of this is provide some basic demographic data that was presented to Avivo on March 12,2018 and incorporates the changes, most notably a distinict 18 to 26 age group, from that meeting

The sources are the custom made demographics data and the discharge portion of the Daanes file.  
For the demographics data, I calculated age at time of the discharge.  For race, I broke out the original race column and created 4 separate columns to make the data tidy (one variable per column).  Please not that these are the number of unique patients as identified by Client.Number.  Because there are clients with multiple visits, the number of patients is smaller than the number of visits.

Contents include:

* Previewing the data

* Counts by sex, primary race (race), and age grouping (age)

* Counts and percentages by age/sex, race/sex, age/race

* Success Rate by sex, race, and age

* Highest success rate by age/sex, race/sex, age/race

* Success rate by all 3 demographic categories

```{r "setup", include=FALSE}
require("knitr")
opts_knit$set(root.dir = "C:\\Users\\user\\Desktop\\nc\\stareightytwo\\Avivo")
```

Accessing and setting up the data:
```{r access_data, results='hide'}
demo <- read.csv("AvivoDemographicTablev3.csv",header = TRUE, sep=",")
library(data.table)
library(dplyr)
library(ggplot2)
library(stringr)
opts_knit$set()
```

Previewing the data:
```{r preview_data}
dim(demo)
summary(demo)
str(demo)
```


**Count by Single Variable Graphs**

Count by sex graph:
```{r sex_graph}
ggplot(data=demo, aes(x=Sex)) + 
     geom_bar(mapping= aes(fill=Sex)) +
     ggtitle("Count of Avivo Clients by Sex") +
     geom_text(stat='count',aes(label=..count..),vjust=0)
```

Fairly evenly divided between male and female.  What is notable is the number of NAs (184 or 12%)

Count by primary race graph:
```{r race_graph}
ggplot(data=demo, aes(x=Race1)) + 
     geom_bar(mapping= aes(fill=Race1)) +
     ggtitle("Count of Avivo Clients by Primary Race") +
     geom_text(stat='count',aes(label=..count..),vjust=0) +
     theme(axis.text.x=element_blank(),axis.ticks.x=element_blank()) +
     xlab("Primary Race")
```

Because race for demographic purposes is calculated by primary race, I broke the primary race by the first value in the race column from the Daanes file.  African-Americans make up roughly half of the non-NA population, followed by patients who are White and then Native Americans.  Hispanic or Latino and Asians appear in very small numbers.


For the age grouping, I changed the age grouping to include 18 and 26 and then segmenting the other age groups by 10 year intervals until reaching age 57.  Note age is as of the discharge date.  This was done by changing the factor for age to numeric and using arbitrary cut points, but generally by decade.

```{r setting_age}
demo[4] <- lapply(demo[4], as.numeric)

demo$Age_group <- cut(demo$Age,
           breaks = c(-Inf, 27, 37, 47, 57, Inf),
           labels = c("18 to 26", "27 to 36", "37 to 46", "47 to 56", "57+"),
           right = FALSE)
```

Count by age grouping graph:
```{r age_graph}
ggplot(data=demo, aes(x=Age_group)) + 
     geom_bar(mapping= aes(fill=Age_group)) +
     ggtitle("Count of Avivo Clients by Age Grouping") +
     geom_text(stat='count',aes(label=..count..),vjust=0) +
     theme(axis.text.x=element_blank(),axis.ticks.x=element_blank()) +
     xlab("Age Grouping")
```

The majority of the patients are between 18 and 49 years old.  There is a drop in the 50-59 age range and then it drops significantly at 60.

**Count by 2 Variable Graphs**

With the 2 variable graphs, I need to explain how to read the graphs.  The y-axis values are represented by the height of the bar.  That was done to show scale.  The value is the percentage within the x-axis variable.


Dataframe establishing count by age and sex:
```{r df_age_sex}
agesex <- demo %>%
    group_by(Age_group,Sex) %>%
    summarize(counts = n()) %>%
    mutate(percent_total = round(counts / sum(counts) * 100,1))
```


Count by age and sex graph:
```{r age_sex_graph}
ggplot(data=agesex, aes(x=Age_group, y=counts, fill=Sex))+
    geom_bar(stat="identity",position = 'dodge') +
    geom_text(aes(label=percent_total),color='black',position = position_dodge(width = 1), vjust = -0.5, size = 3) +
    ggtitle("Percent of Avivo Clients per Age Grouping by Sex")+
    xlab("Age Grouping")
```

Almost 60% of the clients aged 18 to 36 are female, it is even distribution from 37 to 46, and roughly 70% of the clients aged 47+ are male.  In conclusion, there is a shift from female to male as the clientele is older.


dataframe establishing count by primary race and sex:
```{r df_race_sex}
primaryrace_sex <- demo %>%
    group_by(Race1,Sex) %>%
    summarize(counts = n()) %>%
    mutate(percent_total = round(counts / sum(counts) * 100,1))
```

Percent by age and sex graph:

```{r race_sex_graph}
ggplot(data=primaryrace_sex, aes(x=Race1, y=counts, fill=Sex))+
    geom_bar(stat="identity",position = 'dodge') +
    geom_text(aes(label=percent_total),color='black',position = position_dodge(width = 1), vjust = -0.5, size = 3) +
    ggtitle("Percent of Avivo Clients per Primary Race by Sex")+
    xlab("Age Grouping") + ylab("Counts") + 
    scale_x_discrete(labels = function(x) str_wrap(x, width = 10))
```

Almost 2/3 of the African-American patients are male, while over 80% of the Native American patients are female.


Dataframe establishing count by primary race and age group:
```{r df_race_age}
primaryrace_agegroup <- demo %>%
  group_by(Race1,Age_group) %>%
  filter(Race1 %in% c("American Indian or Alaska Native","Black or African American","White")) %>%
  summarize(counts = n()) %>%
  mutate(percent_total = round(counts / sum(counts) * 100,1))

```

Percent by race and age graph

```{r race_age_graph}
ggplot(data=primaryrace_agegroup, aes(x=Race1, y=counts, fill=Age_group))+
  geom_bar(stat="identity",position = 'dodge') +
  geom_text(aes(label=percent_total),color='black',position = position_dodge(width = 1), vjust = -0.5, size = 3) +
  ggtitle("Percent of Avivo Clients per Primary Race by Age Group")+
  xlab("Age Grouping") + ylab("Counts") + 
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10))
```

Given that there were 3 main primary races in the population, the data has been filtered to American Indian or Alaska Native, African-American and White.  The age distribution for American Indian and White show that over 50% of the clients are under age 37 while for African-Americans, 65% of the clients are aged 37 and over.

**Success by single demographic variable graphs**

Before getting into these graphs, the count now shifts from the number of patients to the number of visits.  As a result the demographic and discharge data needs to be merged.

Taking only the discharge data because it this point it the same as the admission data and then merging it with the demographic data:

```{r merge}
avivo <- read.csv("admission_discharge_demographic.csv")
avivo_disc <- subset(avivo, avivo$Form.Flag == "discharge")
avivo_disc_demo <- merge(avivo_disc, demo, by="Client.Number")
```

Setting up variable indicating completion for Completed/Transferred:
``` {r completed value}
avivo_disc_demo$Completed_Program <- 0
avivo_disc_demo$Completed_Program <- ifelse (avivo_disc_demo$Reason.for.Discharge %in% c("Completed program","Transferred to other program"), 1, 0)
avivo_disc_demo$Completed_Program <- as.factor(avivo_disc_demo$Completed_Program)
```


Dataframe establishing success rate by sex (gender)

```{r success_df_sex}
df_success_sex <- avivo_disc_demo %>%
  group_by(Sex.y, Completed_Program) %>%
  summarize(counts = n()) %>% 
  mutate(percent_total = round(counts / sum(counts) * 100,1))
```

Success Rate by sex graph:

```{r success_sex_graph}
ggplot(data = df_success_sex, aes(x = Sex.y, y = counts, fill = Completed_Program))+ 
  geom_bar(position = "dodge", stat = "identity") + 
  geom_text(aes(label=percent_total),color='black',position = position_dodge(width = 1), vjust = -0.5, size = 3) +
  xlab("Sex") + ylab("Count") +
  ggtitle("Success Rate by Sex") + 
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10))
```

Females have a higher success rate than males.  Note that the NA success rate is higher than both.

Dataframe establishing success rate by primary race.  It is filtered by the three primary races.

```{r success_df_race}
df_success_race<- avivo_disc_demo %>%
  group_by(Race1, Completed_Program) %>%
  filter(Race1 %in% c("American Indian or Alaska Native","Black or African American","White")) %>%
  summarize(counts = n()) %>% 
  mutate(percent_total = round(counts / sum(counts) * 100,1))

```

Success rate by primary race graph:

```{r success_race_graph}
ggplot(data = df_success_race, aes(x = Race1, y = counts, fill = Completed_Program))+ 
  geom_bar(position = "dodge", stat = "identity") + 
  geom_text(aes(label=percent_total),color='black',position = position_dodge(width = 1), vjust = -0.5, size = 3) +
  xlab("Primary Race") + ylab("Percent") +
  ggtitle("Success Rate by Primary Race") + 
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10)) +
  theme(axis.text=element_text(size=8))
```

African Americans have the highest completion rates.  They are 5 points higher over Whites and 11 points over American Indians.

Dataframe establishing success rate by age group:
```{r success_df_age}
df_success_age<- avivo_disc_demo %>%
  group_by(Age_group, Completed_Program) %>%
  summarize(counts = n()) %>% 
  mutate(percent_total = round(counts / sum(counts) * 100,1))
```

Success rate by age group graph:

```{r success_age_graph}
ggplot(data = df_success_age, aes(x = Age_group, y = counts, fill = Completed_Program))+ 
  geom_bar(position = "dodge", stat = "identity") + 
  geom_text(aes(label=percent_total),color='black',position = position_dodge(width = 1), vjust = -0.5, size = 3) +
  xlab("Age Group") + ylab("Count") +
  ggtitle("Success Rate by Age Group") + 
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10))
```

Basically, the older the client, the more likely the client will succeed with a noticeable jump starting at age 47.

**Success by 2 demographic variables graphs**

The part shows the success rate by 2 variables but rather than show all values, the highest are shown.  Size is also not taken into account.

Dataframes establishing success rate by age group and sex:
```{r df_success_age_sex}
df_success_agesex <- avivo_disc_demo %>%
  group_by(Age_group, Sex.y, Completed_Program) %>%
  summarize(counts = n()) %>% 
  mutate(percent_total = round(counts / sum(counts) * 100,1))

df_success_agesex$agesex <- paste(df_success_agesex$Age_group, df_success_agesex$Sex.y, sep="_")
```
Filtering out the unknowns:
``` {r filter_unknowns}
df_success_agesex_comp <- subset(df_success_agesex, Completed_Program == 1, Sex.y != "Unknown")
```

Success rate by age/sex group graph but only showing the success rate and ranked:
```{r success_age_sex_graph}
ggplot(data = df_success_agesex_comp, aes(reorder(x = agesex,percent_total), y = percent_total, fill = percent_total))+ 
  geom_bar(position = "dodge", stat = "identity") + 
  geom_text(aes(label=percent_total),color='black',position = position_dodge(width = 1), hjust = 0, size = 3) +
  xlab("Age Group") + ylab("Count") +
  ggtitle("Success Rate by Age/Sex Group, Overall = 43%") + 
  coord_flip() +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10))+
  theme(axis.text=element_text(size=8))
```

Generally what is shows is that older patients are doing better.  Male patients aged 18 to 26, 27 to 36, and 37 to 46 comprise of the bottom 3 groups. 

Dataframes establishing success rate by primary race and sex.  
```{r success_df_race_sex}
df_success_racesex <- avivo_disc_demo %>%
  group_by(Race1, Sex.y, Completed_Program) %>%
  filter(Race1 %in% c("American Indian or Alaska Native","Black or African American","White")) %>%
  summarize(counts = n()) %>% 
  mutate(percent_total = round(counts / sum(counts) * 100,1))

df_success_racesex$racesex <- paste(df_success_racesex$Race1, df_success_racesex$Sex.y, sep="_")
```

Filtering out the unknowns:
```{r filter_unknowns2}
df_success_racesex_comp <- subset(df_success_racesex, Completed_Program == 1, Sex.y != "Unknown")
```

Success rate by primary race/sex graph but only showing the success rate and ranked:
``` {r success_race_sex_graph}
ggplot(data = df_success_racesex_comp, aes(reorder(x = racesex,percent_total), y = percent_total, fill = percent_total))+ 
  geom_bar(position = "dodge", stat = "identity") + 
  geom_text(aes(label=percent_total),color='black',position = position_dodge(width = 1), hjust = 0, size = 3) +
  xlab("Primary Race / Sex") + ylab("Percent") +
  ggtitle("Success Rate by Primary Race/Sex Group, Overall = 43%") + 
  coord_flip() +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10))+
  theme(axis.text=element_text(size=8))
```

African American females have the highest success rates.  Native American males have the lowest rates.

Dataframes establishing success rate by primary race and sex:
```{r success_df_race_age}
df_success_raceage <- avivo_disc_demo %>%
  group_by(Race1, Age_group, Completed_Program) %>%
  filter(Race1 %in% c("American Indian or Alaska Native","Black or African American","White")) %>%
  summarize(counts = n()) %>% 
  mutate(percent_total = round(counts / sum(counts) * 100,1))

df_success_raceage$raceage <- paste(df_success_raceage$Race1, df_success_raceage$Age_group, sep="_")

df_success_raceage_comp <- subset(df_success_raceage, Completed_Program == 1)
```

Success rate by primary race/age group graph but only showing the success rate and ranked:

```{r success_race_age_graph}
ggplot(data = df_success_raceage_comp, aes(reorder(x = raceage,percent_total), y = percent_total, fill = percent_total))+ 
  geom_bar(position = "dodge", stat = "identity") + 
  geom_text(aes(label=percent_total),color='black',position = position_dodge(width = 1), hjust = 0, size = 2) +
  xlab("Primary Race / Age") + ylab("Percent") +
  ggtitle("Success Rate by Primary Race/Age Group, Overall = 43%") + 
  coord_flip() +
  theme(axis.text=element_text(size=6))
```

**Success by all demographic variables graph**

Using all 3 demographic categories but filtering by the 3 primary races and excluding the aged 57+ clientele, and finally setting the number of completed patients by 15+.
```{r success_df_all}
df_success_all <- avivo_disc_demo %>%
  group_by(Race1, Sex.y, Age_group, Completed_Program) %>%
  filter(Race1 %in% c("American Indian or Alaska Native","Black or African American","White"),
         Age_group %in% c("18 to 26","27 to 36","37 to 46","47 to 56")) %>%
  summarize(counts = n()) %>% 
  mutate(percent_total = round(counts / sum(counts) * 100,1))

df_success_all$all <- paste(df_success_all$Race1,df_success_all$Sex.y ,df_success_all$Age_group, sep="_")

df_success_all_comp <- subset(df_success_all, Completed_Program == 1)

counts_not_small <- df_success_all_comp$counts > 14

df_success_all_comp_not_small <- df_success_all_comp[counts_not_small,]
```

Success rate by all 3 demographic categories but only showing the success rate and ranked:
```{r success_all_graph}
ggplot(data = df_success_all_comp_not_small, aes(reorder(x = all, percent_total), y = percent_total, fill = percent_total))+ 
  geom_bar(position = "dodge", stat = "identity") + 
  geom_text(aes(label=percent_total),color='black',position = position_dodge(width = 1), hjust = 0, size = 2) +
  xlab("Primary Race / Age") + ylab("Percent") +
  ggtitle("Success Rate by all Factors, Completed Clients >= 15") + 
  coord_flip() +
  theme(axis.text=element_text(size=6))
```

First, the age/gender/race categories where the number of completed clients below 15 were omitted.  There is one population that doesn't fit the overall story, which is that the younger age groups, 18 to 26 and 27 to 36, has the lowest success rate.  The African-American female population in those age groups has success rates of 53.3% (24 out of 45) and 48.9% (44 out of 90), which are both above average.  Are possible factors, the type of program or the possibility of the impact of children?





